---
title: "Class 11 Lab"
author: "Dennis Kim"
format: gfm
editor: visual
---

#Population Scale Analysis

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes

```{r}
pop.data <- read.table("rs8067378_ENSG00000172057.6.txt")
summary(pop.data)
```

There are 462 samples in total

```{r}
table(pop.data$geno)
```

There are 108 homozygous A, 233 heterozygous A/G, and 121 homozygous G.

```{r}
library(ggplot2)
```

```{r}
boxplot <- ggplot(pop.data) + aes(geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
boxplot
```

```{r}
summary(boxplot$data)
```

AA has a median of 30, AG has a median of 25 and GG has a median of 20

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Time for a boxplot

```{r}
library(ggplot2)
```

```{r}
boxplot <- ggplot(pop.data) + aes(geno, exp, fill=geno) + geom_boxplot(notch=TRUE)
boxplot
```

The expression of AA is much higher than the expression of GG, indicating that the SNP affects the expression of ORMDL3
