---
title: "HW6"
author: "Dennis Kim"
format: pdf
---

# N.B. Homework with scoring rubric.
## Q6. How would you generalize the original code above to work with any set of input protein structures?

```{r}
library(bio3d)
#must load `bio3d`otherwise you cannot access the data on an empty Rstudio tab
```

## Original Code to use for analysis
```{r}
protein.and.drug.analysis <- function(protein.list) {
for (x in protein.list) {
  #read the pdb and shortcut it
  pdb <-read.pdb(x)
  # search for the specifications
  s1.chainA <- trim.pdb(pdb, chain = "A", elety = "CA")
  s1.b <- s1.chainA$atom$b
  #plot
  plotb3(s1.b, sse = s1.chainA, type = "l", ylab = "Bfactor")
}
}
#need to set x for the sources
  x <- c("4AKE", "1AKE", "1E4Y")
#display the function
protein.and.drug.analysis(x)
```

# Comments on what are the inputs to the function.
The input of this function are protein data bank (pdb) structures of a protein either with or without a drug.

# What the function does and how to use it.
The function reads through a list of protein codes, pulling their pdb files and stores them in the new variable “pdb”. We trim the data for the data we want to analyze using the `trim.pdb()` function to pull data for an Alpha Chain of a protein and store in "s1.chainA". We filter for B factors in the atom column of the data. The plot is generated for B-Factor (atomic mobility) vs the residues of a protein structure with/without a drug.

# What is the output of the function.
The output of the function is a plot of B-factor vs protein alpha chain residues of all the proteins in the list of proteins that matches the specifications we called for.
